\LARGE \bf Planning with Attitude

### 2020-09-09 paper/iros_abstract.tex
IEEE formatted paper

% Formatting
\usepackage{xcolor}
\usepackage{bm}  % for bold symbols 
\usepackage{booktabs}  % better tables
\usepackage{pifont}  % for x mark
\usepackage{graphicx}

@@

@@

        \subsubsection{Error Quadratic}
            Rather than simple subtraction, we can use a quadratic function on the
            three-parameter error state \eqref{eq:state_error}:
            \begin{equation} \label{eq:error_quadratic}
                J_\text{err} = \half \phi^T Q \phi 
                = \half \left(\varphi^{-1}(\delta q)\right)^T Q 
                        \left(\varphi^{-1}(\delta q)\right).
            \end{equation}
            where $\delta q = L(q_g)^T q$, and $\phi = \varphi{\delta q}$. The gradient
            and Hessian of \eqref{eq:error_quadratic} are
            \begin{equation}
                \nabla J_\text{err }= \phi^T Q  D(\delta q)  G(\delta q)
            \end{equation}
            \begin{equation}
                \begin{aligned}
                     \nabla^2 J_\text{err} &= 
                        G(\delta q)^T \! \left(
                        D(\delta q)^T Q D(\delta q) + \nabla D \right) G(\delta q) 
                      + I_3 (\phi^T Q  D(\delta q)) \delta q %\\
                \end{aligned}
            \end{equation}
            where, for the Cayley map,
            \begin{equation}
                D(q) = \pdv{\varphi^{-1}}{q} = -\frac{1}{q_s^2}\begin{bmatrix}
                    q_v \;\; & -\frac{1}{q_s} I_3
                \end{bmatrix}
            \end{equation}
            \begin{equation}
                \nabla D = \pdv{q}(D(q)^T Q \phi) 
                = -\frac{1}{q_s^2} \begin{bmatrix} 
                    -2 \frac{q_v}{q_s}^T Q \phi & \phi^T Q \\
                                   Q \phi & 0 \\
                \end{bmatrix}.
            \end{equation}

>>>>

% Formatting
\usepackage{xcolor}
\usepackage{bm}  % for bold symbols 
\usepackage{booktabs}  % better tables
\usepackage{pifont}  % for x mark
\usepackage{graphicx}
\usepackage{hyperref}

@@

\begin{abstract}
Planning and controlling trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces an accessible and unified approach for tracking control
and optimization-based planning on the space of rotations based on vector calculus and
linear algebra. The methodology is used to derive an extension of the Linear-Quadratic
Regulator (LQR) to systems with arbitrary attitudes, which we call Multiplicative LQR
(MLQR). We compare MLQR to a state-of-the-art geometric tracking controller designed
specifically for quadrotors and demonstrate superior tracking and robustness. An
iterative variant of MLQR is compared to existing methods on useveral constrained
trajectory optimization problems, again demonstrating improved runtime performance. %to
optimize trajectories for a variety of robotic systems, we provide benchmark comparisons
between several of the most common attitude representations used for motion planning and
control and find that the combination of unit quaternion state variables with Rodrigues
parameters to represent attitude errors provides an excellent combination of performance
and algorithmic simplicity.
\keywords{
    motion planning and control, quaternions, optimal control, linear quadratic regulator
}
\end{abstract}

@@

        \subsubsection{Error Quadratic}
            Rather than simple subtraction, we can use a quadratic function on the
            three-parameter error state \eqref{eq:state_error}:
            \begin{equation} \label{eq:error_quadratic}
                J_\text{err} = \half \phi^T Q \phi 
                = \half \left(\varphi^{-1}(\delta q)\right)^T Q 
                        \left(\varphi^{-1}(\delta q)\right).
            \end{equation}
            where $\delta q = L(q_g)^T q$, and $\phi = \varphi{\delta q}$. The gradient
            and Hessian of \eqref{eq:error_quadratic} are
            \begin{equation}
                \nabla J_\text{err }= \phi^T Q  D(\delta q)  G(\delta q)
            \end{equation}
            \begin{multline}
                \nabla^2 J_\text{err} = 
                    G(\delta q)^T \! \left(
                    D(\delta q)^T Q D(\delta q) + \nabla D \right) G(\delta q)  \\
                    + I_3 (\phi^T Q  D(\delta q)) \delta q %\\
            \end{multline}
            where, for the Cayley map,
            \begin{equation}
                D(q) = \pdv{\varphi^{-1}}{q} = -\frac{1}{q_s^2}\begin{bmatrix}
                    q_v \;\; & -\frac{1}{q_s} I_3
                \end{bmatrix}
            \end{equation}
            \begin{equation}
                \nabla D = \pdv{q}(D(q)^T Q \phi) 
                = -\frac{1}{q_s^2} \begin{bmatrix} 
                    -2 \frac{q_v}{q_s}^T Q \phi & \phi^T Q \\
                                   Q \phi & 0 \\
                \end{bmatrix}.
            \end{equation}

        \subsubsection{Geodesic Distance}
            Alternatively, we can use the geodesic distance between two quaternions
            \cite{Kuffner2004},
            \begin{equation} \label{eq:quat_geodesic}
                J_\text{geo} = (1-\abs{q_g^T q}) ,
            \end{equation}
            whose gradient and Hessian are,
            \begin{align}
                \nabla J_\text{geo} &= \pm q_g^T G(q) \\
                \nabla^2 J_\text{geo} &= \pm I_3 q_g^T q ,
            \end{align}
            where the sign of the Hessian corresponds to the sign of $q_d^T q$.

\section{Experiments} \label{sec:experiments}
    Code for all experiments is available on 
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

    \subsection{Trajectory Optimization}
        In this section we present several trajectory optimization problems solved using
        our modified version of ALTRO. The two quadrotor examples use 101 knot points for
        5 second trajectories, and the airplane example uses 51 knot points for a 1.25
        second trajectory. All results were run on a desktop computer with an Intel
        i9-9900 3.10 GHz processor with 32GB of memory. Timing results for all three
        experiments are given in Fig. \ref{fig:timing_chart}.

        In all examples, MLQR is implemented with a unit quaternion attitude state and
        Rodrigues parameters as the error state. The first example compares the two cost
        functions given in Sec. \ref{sec:cost_functions}, while the last two use the
        geodesic quaternion cost and compare iterative MLQR (iMLQR) from Sec.
        \ref{sec:MLQR} to iLQR using roll-pitch-yaw Euler angles following the convention
        from \cite{michael2010grasp}, and quaternions (treating them na\"ively as
        vectors). All examples re-normalize the quaternion in the dynamics function and
        use a third-order explicit Runge-Kutta integrator.

        \begin{figure}
            \centering
            \includegraphics[height=4cm, width=8cm]{figures/timing_chart.tikz}
            \caption{Timing results for the airplane barrell roll, quadrotor
                flip, and quadrotor zig-zag examples using iLQR with
                roll-pitch-yaw Euler angles (RPY), quaternions, and iMLQR. The
                timing result for the Quadrotor flip with Euler angles is omitted
                since it failed to converge.
            }
            \label{fig:timing_chart}
        \end{figure}

        \subsubsection{Quadrotor Zig-Zag}
            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm]{figures/zig_zag_tracking.png}
                \caption{Zig-zag trajectory. The black quadrotors indicate
                    the reference trajectory generated by iMLQR. The yellow
                    quadrotors track the reference trajectory using time-varying
                    MLQR, while the red uses the HFCA controller from
                    \cite{watterson2020control}.
                }
                \label{fig:zig-zag}
            \end{figure}    
            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm,trim={0 3cm 0 0},clip]{figures/quad_flip.png}
                \caption{Snapshots of the quadrotor flip trajectory. The
                    red-colored quadrotors represent the state near t=0 s and the
                    green-colored quadrotors represent the state near t=5.0 s
                }
                \label{fig:quad_flip}
            \end{figure}

        We set up a fairly simple trajectory optimization problem to achieve a
        ``zig-zag'' pattern, as shown in Fig. \ref{fig:zig-zag} using cost funtions of
        the form
        \begin{multline}
                  \half \sum_{k \in \mathcal{N}} \ell(x_k, x_\text{nom}, Q_\text{nom}) 
                + \half \sum_{k \in \mathcal{W}} \ell(x_k, \bar{x}_k, Q_w) \\
                + \half \sum_{k=1} u_k^T R u_k
        \end{multline}
        with $\mathcal{W} = \{33,66,101\}$, $\mathcal{N} = \{1:101\} \setminus
        \mathcal{W}$, and the waypoint positions were $(10,0,1)$,$(-10,0,1)$ and
        $(0,10,0)$. We compared the two cost functions from Sec.
        \ref{sec:cost_functions}, using a standard quadratic penalty on the
        non-quaternion states.
        \begin{align}
            \ell_\text{err}(x, \bar{x}, Q) &= \delta x^T Q \delta x \\ 
            \ell_\text{geo}(x, \bar{x}, Q) &= (x-\bar{x})^T \bar{Q} (x-\bar{x}) + w (1
                \pm \bar{q}^T q)
        \end{align}
        where $\bar{Q} = \text{diag}(Q_r, \vec{0}_4, Q_v, Q_\omega)$, where
        $Q_r,Q_v,Q_\omega$ are the weights of $Q$ for position, and linear and angular
        velocity.
        We used the following weighting matrices: 
        $Q_\text{nom} = \text{diag}(10^{-5} I_6,10^{-3} I_6)$, 
        $Q_w = \text{diag}(10^3 I_3,I_9)$, 
        $Q_N = \text{diag}(10 I_3, 100 I_3, 10 I_6)$.
        For the geodesic weights $w$, we used 0, 1, and 10 for the nominal, waypoint, and
        terminal cost functions, respectively.

        The geodesic cost function performed significantly better than the error
        quadratic, converging in 10 ms and 13 iterations, compared to 53 ms in 29
        iterations. In addition to better convergence behavior, the geodesic cost is
        about half as expensive to compute, which is unsurprising when comparing the
        gradients and Hessians in Sec. \ref{sec:cost_functions}.

        \subsubsection{Quadrotor Flip}
	    Similar to the previous example, we use a set of waypoints to encourage the
	    quadrotor to do a flip. The cost weights for this example were
	    \begin{align*}
	        Q_\text{nom} &= \text{diag}(10^{-2},10^{-2},5 \!\times\! 10^{-2}, 10^{-3} I_3,
    	        10^{-3} I_3, 10^{-2} I_3) \\
	        Q_w &= \text{diag}(10^3, 10, 10^3, 5000 I_3, I_3, 10 I_3) \\
	        Q_N &= \text{diag}(10 I_3, 100 I_3, 10 I_6)  
	    \end{align*}
	    Four intermediary knotpoints were used to encourage the quadrotor to be at angles
	    of \ang{90}, \ang{180}, \ang{270}, and \ang{360}. The quadrotor was constrained
	    to stay above the floor and move to a goal state 2 meters away in the $+y$
	    direction. The solver is initialized with a dynamically infeasible trajectory
	    that linearly interpolates between the initial and final states, rotating the
	    quad around the x-axis a full \ang{360}.

	    Figure \ref{fig:quad_flip} shows snapshots of the trajectory as generated using
	    iMLQR within the ALTRO solver to handle constraints. The trajectory generated by
	    iLQR with quaternions and quadratic costs was successful but had a slightly
        different trajectory. iLQR with Euler angles failed to solve the problem.

        \subsection{Airplane Barrel Roll}
        An airplane model with aerodynamic coefficients fit from real wind-tunnel data is
        tasked to do a barrel roll by setting a high terminal cost for being upside-down,
        see Fig. \ref{fig:barrellroll}. The solver is initialized with level flight trim
        conditions. For both the quadrotor flip and the airplane barrel roll, iMLQR
        converged faster than the pure quaternion version. Despite the extra matrix
        multiplications and extra Hessian term, it also was faster per iteration than its
        iLQR counterpart. For these highly aerobatic maneuvers, we achieve, as expected,
        better performance by correctly leveraging the structure of the rotation group
        during the optimization routine.

        \begin{figure}[h]
            \centering
            \includegraphics[height=3.5cm]{figures/barrellroll.png}
            \caption{Barrel roll trajectory computed by iterative MLQR using a terminal cost to encourage an upside-down attitude.}
            \label{fig:barrellroll}
        \end{figure}

	    \begin{figure}[h]
	        \centering
	        \includegraphics[width=\columnwidth,height=4.5cm]{figures/tracking_err.tikz}
	        \caption{Tracking error over time for the zig-zag trajectory using MLQR and the HFCA geometric controller. Band-limited white noise wrenches of variance $w$ are added to the continuous dynamics. }
	        \label{fig:tracking_err}
	    \end{figure}
        \vspace{-20pt}

	\subsection{Tracking Control}
	    This section provides examples using MLQR for tracking and stabilizing control
	    for a quadrotor. We compare our MLQR controller to HFCA, a state-of-the-art
	    geometric tracking controller \cite{watterson2020control} based on the Hopf
	    Fibration.

	    \subsubsection{MLQR Stabilization}
	    To test robustness and the domain of attraction of the MLQR controller, a
	    quadrotor is linearized about hover at the origin and initialized from states
	    uniformly sampled in $[-1,1]$ for position, $[-5,5]$ for linear and angular
	    velocities, and uniformly over the entire space of rotations. The success rates
	    for 2000 trials are given in Table \ref{tab:mlqr_success}, where success is
	    defined as the terminal error (calculated using \eqref{eq:state_error}) being
	    within a unit ball of radius 0.1 after 10.0 seconds. To further analyze the
	    basins of attraction, we uniformly sampled over only roll (which goes singular at
	    \ang{90} for our Euler angle representation), and yaw. The results are shown in
	    Fig. \ref{fig:mlqr_basin}, where the plotted contours are the convex hull of the
	    successfully stabilized initial conditions.

	    MLQR slightly exceeds the robustness of the HFCA geometric controller, while also
	    being more generally applicable, easier to tune, and easier to implement. It is
	    also noteworthy that na\"ively using LQR with quaternions (incorrectly treating
	    them as vectors) results in poorer performance than Euler angles, clearly
	    demonstrating the value of the methodology presented in the current paper.

	    \subsubsection{MLQR Tracking}
	    Time-varying MLQR is used to track the zig-zag trajectory generated in the
	    previous section. %The controller calculated the gains at $0.01 s$ intervals, and
	    used zero-order-hold between intervals. To test tracking performance, zero-mean
	    Gaussian noise was added as force and moments applied to the quadrotor.

	    The trajectories and tracking error for MLQR and HFCA are shown in Figs.
	    \ref{fig:zig-zag}, and \ref{fig:tracking_err}. While the MLQR controller
	    dramatically out-performs HFCA, we found no significant difference in performance
	    between MLQR and LQR using either roll-pitch-yaw Euler angles or a quaternion.
	    Since the tracking controller only deals with relatively small state and attitude
	    errors, this is unsurprising. Additionally, we found the MLQR controller ran
        about 20x faster than the HFCA controller.

	    \begin{figure*}[ht]
	        \begin{minipage}{0.60\linewidth}
                \centering
                \includegraphics[height=6.6cm,width=\textwidth]{figures/mlqr_basin.tikz}
                \caption{Basins of attraction of 4 stabilizing controllers for yaw and roll, calculated from 2000 sample points (shown in black). All points within each contour were successfully stabilized by the controller.}
                \label{fig:mlqr_basin}
            \end{minipage}
	        \hspace{0.05\linewidth}
            \begin{minipage}{0.34\linewidth}
                \centering
                \begin{tabular}{cc}
                    \toprule
                    \textbf{Method} & \textbf{Success} \\
                    \midrule
                    Quat & 61\% \\
                    RPY  & 80\% \\
                    HFCA & 97\% \\
                    MLQR & 99\% \\
                    \bottomrule
                \end{tabular}
                \caption{Success rate for Monte-Carlo analysis of quadrotor stabilization for 2000 initial conditions uniformly over position, orientation, and linear and angular velocities.}
                \label{tab:mlqr_success}
            \end{minipage}
        \end{figure*}

\section{Conclusions} \label{sec:conclusion}
    We have presented a general, unified method for planning and control on rigid-body
    systems with arbitrary attitude using standard linear algebra and vector calculus. By
    applying these methods to LQR to correctly account for the group structure of
    rotations, we have matched or exceeded the performance of a state-of-the-art
    geometric controller designed specifically for quadrotors, while being more general,
    requiring less system-dependent tuning, having less computational overhead, and being
    easier to implement.

    We have also demonstrated a straight-forward way to adapt nonlinear trajectory
    optimization techniques to work with quaternion-valued states. For these problems, we
    found the geodesic distance between quaternions \eqref{eq:quat_geodesic} to be
    computationally efficient and provide excellent convergence in practice. We recommend
    the use of unit quaternions within planning and control algorithms for their
    simplicity, computational efficiency, and lack of singularities. We further recommend
    the use of Rodrigues parameters, or the Cayley map, as the quaternion error state.

    Future areas of research include the application of these methodologies to more
    complex multi-body floating-base robots, such as humanoids and quadrupeds, as well as
    more in-depth analysis of the convergence behavior of the algorithms proposed in the
    current work.

\paragraph{Acknowledgements}
This material is based upon work supported by the National Science Foundation Graduate
Research Fellowship Program under Grant No. DGE-1656518. Any opinions, findings, and
conclusions or recommendations expressed in this material are those of the author(s) and
do not necessarily reflect the views of the National Science Foundation

\printbibliography
